### 2016-08-22 latex/input/conclusion.tex
updated results in paper, edited plots, wrote script to calculate AUC-ROC

We have demonstrated an effective and efficient computational pipeline for classifying patient phenotype based on metagenomic data. We have demonstrated that even relatively simple de novo assembly and clustering methods, when used within this pipeline, lead to significantly better performance results than the standard classifier used in the original Metagenome-Wide Association Study. We have shown how to infer the most important OTUs in the disease pathology by using the SVM decision boundary and discussed the clinical importance of this ability. More generally, we have shown the effectiveness of Multiple Instance Learning methods within metagenomics and phenotype prediction, particularly Distance-based Bag of Words methods. CAMIL is a relatively simple and easy to implement pipeline that has both shown strong results and significant potential for even further improvement. Future work could revolve around improving individual parts of the pipeline, such as better assembly and clustering methods, application of different multiple instance learning methods (other than D-BoW), and further attempts to generate more specific instance level or cluster level information and validate that information against the known pathology of various diseases.

>>>>

We have demonstrated an effective and efficient computational pipeline for classifying patient phenotype based on metagenomic data. We have demonstrated that even relatively simple de novo assembly and clustering methods, when used within this pipeline, lead to significantly better performance results than the standard classifier used in the original Metagenome-Wide Association Study. We have shown how to infer the most important OTUs in the disease pathology by using the SVM decision boundary and discussed the clinical importance of this ability. More generally, we have shown the effectiveness of Multiple Instance Learning methods within metagenomics and phenotype prediction, particularly Bag of Words methods. CAMIL is a relatively simple and easy to implement pipeline that has both shown strong results and significant potential for even further improvement. Future work could revolve around improving individual parts of the pipeline, such as better assembly and clustering methods, application of different multiple instance learning methods (other than Bag of Words), and further attempts to generate more specific instance level or cluster level information and validate that information against the known pathology of various diseases.

### 2016-08-22 latex/input/experiments.tex
updated results in paper, edited plots, wrote script to calculate AUC-ROC

The 218 publicly-available and labeled patients were downloaded from NCBI and converted to FASTQ format using the SRA toolkit, and labeled according to the labels on EBI, as described in the Materials section. The data was split evenly into training and test sets, with each containing 109 patients. The training set had 54 T2D patients and 55 control patients, while the test set had 53 T2D patients and 56 control patients. Each read was assembled with SOAPdenovo2, with the k-mer length set to 51, reads cut off after 100 base pairs (original length of 180 base pairs), and the average insert size set to 350 in accordance with the reported average insert size reported in the MGWAS study \cite{qin041012}. Assembly for each file took 7-30 minutes, depending on the file size. The assembly of the patient files is embarrassingly parallel, so the assembly of each file could be done at the same time. Combining the files into one file took 12 minutes. Assembly was used for all of the classification methods tested, because the combining of individual reads and reduction in total data size made classification feasible.

@@

This section provides an overview of the different methods that we compared our pipeline to. Our pipeline uses clustering, whereas the other methods do not. Instead, those methods directly compare individual instances instead of clusters. In order to do this, sequence reads were represented as counts of k-mers. k-mers are nucleotide strings of length k. Since there are four possible nucleotides, the number of possible k-mers is \(4^k\). For instance, if k=3, there are \(4^3\) = 64 possible k-mers, which are AAA, AAC, AAG, AAT, ACA, ACC, ..., TTT. For a string "ATACGATA", the count for the k-mers is 2 for ATA, 1 for TAC, ACG, CGA, and GAT, and 0 for everything else. We wrote a script to represent the reads as vectors representing the k-mer counts in the string, and these vectors were used by the other methods. From experimental validation, we found a k-mer value of 3 to be the most effective. Converting the reads from their original string form to k-mer vectors took 3 hours, 17 minutes, and 55 seconds.

@@

The methods in the original paper are neither MIL-based, nor are they entirely de novo apart from patient labels. The authors first performed de novo assembly with SOAPdenovo2 \cite{luo12} and then use a tool called MetaGeneMark \cite{zhu10, besemer99} for de novo prediction of genes from the assembled contigs \cite{qin041012}. They then combined these genes with an existing gene catalog, MetaHIT \cite{qin030410}, and carried out taxonomic assignment and functional annotation of the genes using the KEGG \cite{kanehisa00} and eggNOG \cite{powell12} databases, as well as 2,890 other reference genomes \cite{qin041012}. The authors defined gene markers by mapping the sequence reads from the MGWAS dataset to the updated gene catalog. They identified the 50 most important gene markers with the minimum redundancy - maximum relevance (mRMR) \cite{peng05} method, using the ``sideChannelAttack" R package and then used these 50 gene markers for SVM classification of T2D phenotype, using the ``e1071" R package for the SVM \cite{qin041012}. Thus, the method in the original paper first applies de novo assembly and gene prediction methods, but then uses a number of references to identify the gene markers to be used in classification. From their results, the authors generated an Area Under Curve - Receiver Operating Characteristic (AUC-ROC) graph. We manually computed the accuracy and F1 score with an ideally-chosen decision boundary based on the results the authors provided in their supplementary tables.

@@

\begin{table}[h]
\begin{center}
\caption{Comparison of the performance of various methods on the MGWAS dataset. 
\label{tab:comp}}
\begin{tabular}{|c|ccc|}\hline
Method & Accuracy & F1-Score & AUC-ROC\\\hline
MISVM & 50.00 & 50.00 & 50.00\\\hline
sbMIL & 50.00 & 50.00 & 50.00\\\hline
GICF & 60.00 & 71.43 & 66.00\\\hline
mRMR + SVM & 82.61 & 80.02 & 81.00\\\hline 
CAMIL H-BoW & 91.74 & 91.43 & 95.00\\\hline
CAMIL D-BoW & \bf{92.66} & \bf{92.31} & \bf{96.00}\\\hline
\end{tabular}
\end{center}
\end{table}

Table \ref{tab:comp} shows the comparison of classification performance between the various methods described above. CAMIL methods significantly outperform the other methods, with the D-BoW variant of CAMIL slightly outperforming the H-BoW variant.

MISVM and sbMIL had the worst results, faring no better than a random guess. This makes sense, as they make the standard MIL assumption, which doesn't make sense in the context of phenotype prediction, in which even healthy patients can host a small number of pathogens. Additionally, they are instance space methods that do not leverage bag-level information. The performance of these two methods serves to illustrate why many of the classic MIL algorithms with standard assumptions will not be effective in this domain. GICF performs better than MISVM and sbMIL, which makes sense given the fact that it follows the collective assumption. It also has the benefit of calculating instance labels, which we explore further in the next section. However, GICF is still an instance space method, so it makes sense that it performs worse than CAMIL.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.4]{./instance-scatter.png}
\caption{This diagram illustrates the distribution of the instance weights assigned by CAMIL D-BoW. The Y-Axis shows the weights, while the X-Axis shows the ranking of the clusters by weight. Clearly, there are a relatively small number of clusters that have disproportionately large weights, while there is a smaller number of clusters with smaller weights.} \label{instance-scatter}
\end{figure}

One of the benefits of using Multiple Instance Learning methods is that we can attempt to discover instance ``labels". In fact, we did not attempt to apply static, unchanging labels to individual reads or clusters, since organisms are affected by their interactions with each other. For instance, a patient with X amount of microbe A or X amount of microbe B may be healthy, but with X/2 amount of microbe A and X/2 amount of microbe B they may be sick. This is illustrated in Figure \ref{instance-labels}. This example is simple, but it explains why static instance labels are insufficient.

@@

In the original MGWAS paper, the authors identify 50 important gene markers with mRMR that are used for their classifier. Conversely, CAMIL uses all of the data to train and test the classifier, and subsequently identifies significant clusters based on the classification results. We printed the 25 highest-weighted and 25 lowest-weighted clusters, out of 3188 total clusters. The highest-weighted cluster had a weight of 0.012109, while the lowest-weighted cluster had a weight of -0.003668. The average of the 25 highest cluster weights was 0.010402, while it was -0.002753 for the lowest 25. The average of all 3188 cluster weights was 0.001003, and the median was 0.000254. These results indicate that the highest weights were much larger than the lowest weights and skewed the average weight to be much higher than the median weight. Intuitively, this appears to make sense, as there should be a relatively small number of key clusters whose presence is actually indicative of type 2 diabetes, while most other clusters are not particularly relevant in this case and whose weights are just noise. Thus, the weights obtained by this method appear to be plausible. In contrast, the labels obtained by GICF were barely differentiated from each other at all. The instance weights determined by CAMIL are shown in Figure \ref{instance-scatter}.

>>>>

The 367 patient files were downloaded from NCBI and converted to FASTQ format using the SRA toolkit, and labeled according to the Supplementary Tables of the MGWAS paper \cite{qin041012}, as described in the Materials section. Out of the 367 patients, 182 were diabetic and 185 were healthy controls. Each read was assembled with SOAPdenovo2, with the k-mer length set to 51, reads cut off after 100 base pairs (original length of 180 base pairs), and the average insert size set to 350 in accordance with the reported average insert size reported in the MGWAS study \cite{qin041012}. Assembly for each file took 7-30 minutes, depending on the file size. The assembly of the patient files is embarrassingly parallel, so the assembly of each file could be done at the same time. Combining the files into one file took 2 minutes and 35 seconds. Assembly was used for all of the classification methods tested, because the combining of individual reads and reduction in total data size made classification feasible.

@@

This section provides an overview of the different methods that we compared our pipeline to. Our pipeline uses clustering, whereas the other methods do not. Instead, those methods directly compare individual instances instead of clusters. In order to do this, sequence reads were represented as counts of k-mers. k-mers are nucleotide strings of length k. Since there are four possible nucleotides, the number of possible k-mers is \(4^k\). For instance, if k=3, there are \(4^3\) = 64 possible k-mers, which are AAA, AAC, AAG, AAT, ACA, ACC, ..., TTT. For a string "ATACGATA", the count for the k-mers is 2 for ATA, 1 for TAC, ACG, CGA, and GAT, and 0 for everything else. We wrote a script to represent the reads as vectors representing the k-mer counts in the string, and these vectors were used by the other methods. We tested possible k values from 1 to 6; since the number of possible k-mers is exponential in k, higher values for k quickly become impractical in both run time and memory usage. From experimental validation, we found a k-mer value of 3 to be the most effective. Converting the reads from their original string form to k-mer vectors with k=3 took 24 hours, 8 minutes, and 49 seconds.

@@

The methods in the original paper are neither MIL-based, nor are they entirely de novo apart from patient labels. The authors first performed de novo assembly with SOAPdenovo2 \cite{luo12} and then use a tool called MetaGeneMark \cite{zhu10, besemer99} for de novo prediction of genes from the assembled contigs \cite{qin041012}. They then combined these genes with an existing gene catalog, MetaHIT \cite{qin030410}, and carried out taxonomic assignment and functional annotation of the genes using the KEGG \cite{kanehisa00} and eggNOG \cite{powell12} databases, as well as 2,890 other reference genomes \cite{qin041012}. The authors defined gene markers by mapping the sequence reads from the MGWAS dataset to the updated gene catalog. They identified the 50 most important gene markers with the minimum redundancy - maximum relevance (mRMR) \cite{peng05} method, using the ``sideChannelAttack" R package and then used these 50 gene markers for SVM classification of T2D phenotype, using the ``e1071" R package for the SVM \cite{qin041012}. Thus, the method in the original paper first applies de novo assembly and gene prediction methods, but then uses a number of references to identify the gene markers to be used in classification. From their results, the authors generated an Area Under Curve - Receiver Operating Characteristic (AUC-ROC) graph. The authors did not provide a learned decision boundary in their supplementary tables, only the predicted values for each patient, so we manually computed the accuracy and F1 score with an optimally-chosen decision boundary.

@@

\begin{table*}[t]
\begin{center} 
\begin{minipage}{0.4\textwidth}
\caption{Performance on training set.}
\label{tab:train-comp}
\begin{tabular}{|c|ccc|}\hline
Method & Accuracy & F1-Score & AUC-ROC\\\hline
mRMR + SVM & 73.26 & 76.65 & 80.77\\\hline 
CAMIL D-BoW & 90.70 & 90.91 & 97.45\\\hline
CAMIL H-BoW & \bf{94.19} & \bf{93.87} & \bf{97.87}\\\hline
\end{tabular}
\end{minipage}
\begin{minipage}{0.4\textwidth}
\caption{Performance with 23 patient test set.} 
\label{tab:test-comp}
\begin{tabular}{|c|ccc|}\hline
Method & Accuracy & F1-Score & AUC-ROC\\\hline
%MISVM & 50.00 & 50.00 & 50.00\\\hline
%sbMIL & 50.00 & 50.00 & 50.00\\\hline
GICF & 73.91 & 75.00 & 78.03\\\hline
mRMR + SVM & 82.61 & 82.76 & 87.12\\\hline 
CAMIL D-BoW & 91.30 & 92.31 & \bf{100.00}\\\hline
CAMIL H-BoW & \bf{100.00} & \bf{100.00} & \bf{100.00}\\\hline
\end{tabular}
\bigskip
\end{minipage}
%\hfill
\begin{minipage}{0.4\textwidth}
\caption{Performance with even train/test split.} 
\label{tab:even-comp}
\begin{tabular}{|c|ccc|}\hline
Method & Accuracy & F1-Score & AUC-ROC\\\hline
%MISVM & 50.00 & 50.00 & 50.00\\\hline
%sbMIL & 50.00 & 50.00 & 50.00\\\hline
GICF & 63.04 & 68.33 & 66.19\\\hline %59.24,63.05,66.19
CAMIL D-BoW & 86.34 & 87.18 & 95.93\\\hline
CAMIL H-BoW & \bf{90.71} & \bf{89.70} & \bf{97.63}\\\hline
\end{tabular}
\end{minipage}
\begin{minipage}{0.4\textwidth}
\caption{Classification time with even train/test split.} 
\label{tab:time-comp}
\begin{tabular}{|c|c|}\hline
Method & Classification Time\\\hline
%MISVM & 1 hour\\\hline
%sbMIL & 1 hour\\\hline
GICF & 8 hours, 44 minutes, 27 seconds\\\hline
CAMIL D-BoW & \bf{6 minutes, 58 seconds}\\\hline
CAMIL H-BoW & 7 minutes, 56 seconds\\\hline
\end{tabular}
\end{minipage}
\end{center}
\end{table*}

%\begin{table}[t]
%\begin{center}

%\end{center}
%\end{table}

%\begin{table}[t]
%\begin{center}

%\end{center}
%\end{table}

In the original MGWAS paper, the authors use 344 patients as a training set and 23 as a test set. However, in their reported AUC-ROC reflects the performance of of the mRMR + SVM approach on the 344 patients whose data was used to generate the gene markers that were used for classification. Essentially, they tested the performance of their approach on the training data. We thus applied our CAMIL pipeline in the same manner, using the same 344 patients as both the training and test set. The results are shown in Table \ref{tab:train-comp}, which demonstrates that both CAMIL variants significantly outperformed the mRMR + SVM approach in all metrics by about 14-17\%, with the CAMIL H-BoW version performing the best. Since this is not a typical way of assessing classification performance, we ran several other experiments, discussed below.

The MGWAS authors also provided the calculated labels for the 23 patients in the test set in their supplementary tables \cite{qin041012}, although they did not explicitly report the AUC-ROC on the test set. However, based on the reported patient labels, we were able to calculate all of the relevant metrics. Table \ref{tab:test-comp} shows the performance of the various algorithms, including GICF, on the test set. GICF was outperformed by mRMR + SVM, which was in turn outperformed by CAMIL D-BoW, which was in turn outperformed by CAMIL H-BoW, which got perfect results.

Since the train/test split in Table \ref{tab:test-comp} is so imbalanced, we wanted to evaluate CAMIL with more balanced training and test sets. Since this was no longer using the same training set as the original MGWAS paper, we could not evaluate mRMR + SVM. Table \ref{tab:even-comp} shows the performance of CAMIL and GICF with an even split between the training and test sets. CAMIL methods significantly outperform GICF, with the H-BoW variant of CAMIL slightly outperforming the D-BoW variant.

Finally, Table \ref{tab:time-comp} shows the classification time of GICF and CAMIL. mRMR + SVM was not included because the runtime of that method was not reported in the MGWAS paper. CAMIL took significantly less time than GICF for two main reasons: (i) GICF requires representing each read as an array of length 64 (for k-mer length 3), while CAMIL reduces the data size with clustering and feature extraction; (ii) GICF requires expensive pairwise comparisons between each pair of instances in a mini-batch.

MISVM and sbMIL require computing a kernel matrix of size N*N, where N = the number of instances. Since this dataset involved millions of instances, these methods crashed with memory errors, even when we used only 10\% of the reads. We were able to run these methods when using only 1\% of the reads from each patient, and they achieved only 50\% accuracy, no better than a random guess. This makes sense, as they make the standard MIL assumption, which doesn't make sense in the context of phenotype prediction, in which even healthy patients can host a small number of pathogens. Additionally, they are instance space methods that do not leverage bag-level information. The performance of these two methods serves to illustrate why many of the classic MIL algorithms with standard assumptions will not be effective in this domain. GICF performs better than MISVM and sbMIL (albeit with more data), which makes sense given the fact that it follows the collective assumption. It also has the benefit of calculating instance labels, which we explore further in the next section. However, GICF is still an instance space method, so it makes sense that it performs worse than CAMIL.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.4]{./instance-scatter.png}
\caption{This diagram illustrates the distribution of the instance weights assigned by CAMIL H-BoW. The Y-Axis shows the weights, while the X-Axis shows the ranking of the clusters by weight. Clearly, there are a relatively small number of clusters that have disproportionately large weights or small weights, while the vast majority of the 3918 clusters have weights close to 0.} \label{instance-scatter}
\end{figure}

One of the benefits of using Multiple Instance Learning methods is that we can attempt to discover instance ``labels". In fact, we did not attempt to apply static, unchanging labels to individual reads or clusters, since organisms are affected by their interactions with each other. For instance, a patient with X amount of microbe A or X amount of microbe B may be healthy, but with X/2 amount of microbe A and X/2 amount of microbe B they may be sick. This simple example is illustrated in Figure \ref{instance-labels}.

@@

In the original MGWAS paper, the authors identify 50 important gene markers with mRMR that are used for their classifier. Conversely, CAMIL uses all of the data to train and test the classifier, resulting in 3918 clusters, and subsequently identifies significant clusters based on the classification results. Figure \ref{instance-scatter} is a visual display of the cluster weights determined by CAMIL, using the 344 patient training set and 23 patient test set. Clearly, there are a few clusters with disproportionately high or low weights, while most clusters have weights near 0. Concretely, the lowest cluster weight is -0.000333, the highest is 0.000374, the mean is 0.000008, and the median is 0.000006. Intuitively, this appears to make sense, as there should be a relatively small number of key clusters whose presence is actually indicative of type 2 diabetes, while most other clusters are not particularly relevant in this case and whose weights are just noise. Thus, the weights obtained by this method appear to be plausible. In contrast, the labels obtained by GICF were barely differentiated from each other at all.

### 2016-08-22 latex/input/intro.tex
updated results in paper, edited plots, wrote script to calculate AUC-ROC

In this study, we develop an efficient classifier that predicts whether or not a patient has a disease based on their microbiome. We view this as a Multiple Instance Learning (MIL) problem, in which we have several \emph{bags} of instances, and we have labels for the bags, but not for each instance within them. In this case, we have a patient (bag) and a label for each patient (whether or not they have a disease), but no labels for each patient's sequence reads (instances). Specifically, we use a Distance-based Bag of Words (D-BoW) method, discussed further in the Background section, which has been shown to be among the most effective and efficient MIL methods \cite{amores13}. We also show that our pipeline can be used for Histogram-based Bag of Words (H-BoW) methods. MIL has been studied in many contexts, but it has rarely if ever been studied in the context of predicting clinical phenotype based on metagenomic data from the microbiome. However, since datasets in this domain frequently have patient-level labels but almost never have instance-level labels, this is a well-suited domain for multiple instance learning.

>>>>

In this study, we develop an efficient classifier that predicts whether or not a patient has a disease based on their microbiome. We view this as a Multiple Instance Learning (MIL) problem, in which we have several \emph{bags} of instances, and we have labels for the bags, but not for each instance within them. In this case, we have a patient (bag) and a label for each patient (whether or not they have a disease), but no labels for each patient's sequence reads (instances). Specifically, we use Bag of Words (BoW) methods, discussed further in the Background section, which have been shown to be among the most effective and efficient MIL methods \cite{amores13}. MIL has been studied in many contexts, but it has rarely if ever been studied in the context of predicting clinical phenotype based on metagenomic data from the microbiome. However, since datasets in this domain frequently have patient-level labels but almost never have instance-level labels, this is a well-suited domain for multiple instance learning.

### 2016-08-22 latex/input/materials.tex
updated results in paper, edited plots, wrote script to calculate AUC-ROC

We used data from a well-known Metagenome-Wide Association Study by Qin et al. of Type 2 Diabetes (T2D) in Chinese patients \cite{qin041012}. This study was chosen because it is one of the only MGWAS studies that made its data available online and labeled the phenotype of the patients, and is the largest among those studies. The full dataset used in this study contains 368 patients \cite{qin041012}, but only 218 of them were labeled and available online. Each patient file was downloaded from NCBI\footnote{http://trace.ncbi.nlm.nih.gov/Traces/study/?acc=SRP011011}
and converted to FASTQ format using the SRA toolkit\footnote{http://www.ncbi.nlm.nih.gov/Traces/sra/sra.cgi?view=software}.
The labels were found on the EBI website\footnote{https://www.ebi.ac.uk/metagenomics/projects/SRP011011}. The total size of these 218 FASTQ files was 2.3 terabytes, with an average size of 10.82 gigabytes per patient file. Qin et al. develop a simple T2D classifier using a minimum redundancy---maximum relevance (mRMR) method \cite{peng05} for feature selection and an SVM for classification, based on the R packages ``sideChannelAttack" and ``e1071", respectively \cite{qin041012}. They achieved an AUC-ROC of 0.81 by training a classifier on 345 of the patients and using the remaining 23 as a test set \cite{qin041012}. While we did not have access to all of the same patients that they had, we were able to replicate their method and test it on the subset of data that was publicly available. The study also called for more extensive testing of gut microbiota classifiers \cite{qin041012}.

>>>>

We used data from a well-known Metagenome-Wide Association Study by Qin et al. of Type 2 Diabetes (T2D) in Chinese patients \cite{qin041012}. This study was chosen because it is one of the only MGWAS studies that made its data available online and labeled the phenotype of the patients, and is the largest among those studies. The full dataset used in this study contains 367 patients \cite{qin041012}. Each patient file was downloaded from NCBI\footnote{http://trace.ncbi.nlm.nih.gov/Traces/study/?acc=SRP008047, http://trace.ncbi.nlm.nih.gov/Traces/study/?acc=SRP011011}
and converted to FASTQ format using the SRA toolkit\footnote{http://www.ncbi.nlm.nih.gov/Traces/sra/sra.cgi?view=software}.
The labels were found in the paper's Supplementary Tables \cite{qin041012}. The total size of these 367 FASTQ files was 3.29 terabytes, with an average size of 8.97 gigabytes per patient file. Qin et al. develop a simple T2D classifier using a minimum redundancy---maximum relevance (mRMR) method \cite{peng05} for feature selection and an SVM for classification, based on the R packages ``sideChannelAttack" and ``e1071", respectively \cite{qin041012}. They achieved an AUC-ROC of 0.81 by training a classifier on 345 of the patients and using the remaining 23 as a test set \cite{qin041012}. The study also called for more extensive testing of gut microbiota classifiers \cite{qin041012}.

### 2016-08-22 latex/input/methods.tex
updated results in paper, edited plots, wrote script to calculate AUC-ROC

\begin{figure*}[t]
\centering
\includegraphics[scale=0.5]{./mil-metagenomics-pipeline.png}
\caption{This diagram illustrates the entire pipeline. Patient files with FASTQ metagenomic reads are individually assembled using SOAPdenovo2, then combined into one file and clustered with UCLUST. We extract features according to the D-BoW method, and classify the patients using the feature vectors with svm-light.} \label{pipeline}
\end{figure*}

@@

Our proposed pipeline involves a number of steps, which serve a variety of purposes. For each patient file, we assembled the sequence reads, which served the dual purpose of generating larger contigs that contain more functional biological information and reducing the dataset size by discarding reads that could not be assembled. The clustering step assigns the contigs to certain clusters, which represent functionally similar microbes, and thus establish classes of instances that can be used as features for the classifier. We  applied our feature extraction method, which uses the D-BoW method of measuring the closest distance from a patient's reads to each cluster center. Finally, we used an SVM-based classifier to predict patient phenotype, and used several metrics to assess its accuracy. We used the SVM's decision boundary to infer information about which clusters of instances were most or least indicative of the phenotype, discussed further in the Results section. Aside from the patient labels, this process is entirely de novo, and does not consult any external databases. An illustration of this pipeline can be seen in Figure \ref{pipeline} on page \pageref{pipeline}.

@@

Our feature extraction operation followed the D-BoW method. For each patient, we initialized a vector with length equal to the number of clusters. The value for each place in that patient's feature vector would then be set to the closest percentage match (0.4 to 1) of the reads from that patient to the relevant cluster seed. For instance, if there are three clusters of reads, and patient A has the cluster seed for cluster 1, no reads from cluster 2, and a read with 70\% match to cluster 3, their feature vector would be~~A = [1.0~~~0.0~~~0.7]. This is illustrated in Figure \ref{pipeline}. We implemented this feature extraction method in Python.

>>>>

\begin{figure*}[t]
\centering
\includegraphics[scale=0.5]{./mil-metagenomics-pipeline.png}
\caption{This diagram illustrates the entire CAMIL pipeline. Patient files with FASTQ metagenomic reads are individually assembled using SOAPdenovo2, then combined into one file and clustered with UCLUST. We extract features according to either the D-BoW or H-BoW method, and classify the patients using the feature vectors with svm-light.} \label{pipeline}
\end{figure*}

@@

Our proposed pipeline involves a number of steps, which serve a variety of purposes. For each patient file, we assembled the sequence reads, which served the dual purpose of generating larger contigs that contain more functional biological information and reducing the dataset size by discarding reads that could not be assembled. The clustering step assigns the contigs to certain clusters, which represent functionally similar microbes, and thus establish classes of instances that can be used as features for the classifier. We  applied our feature extraction method, which uses the D-BoW method of measuring the closest distance from a patient's reads to each cluster center. Finally, we used an SVM-based classifier to predict patient phenotype, and used several metrics to assess its accuracy. We used the SVM's decision boundary to infer information about which clusters of instances were most or least indicative of the phenotype, discussed further in the Results section. Aside from the patient labels, this process is entirely de novo, and does not consult any external databases. An illustration of the pipeline is shown in Figure \ref{pipeline} on page \pageref{pipeline}.

@@

We test both the Distance-based and Histogram-based feature extraction methods within the CAMIL pipeline. For each patient, we initialized a vector with length equal to the number of clusters. The value for each scalar in that patient's feature vector would then represent either the read that matched most closely to that cluster centroid (Distance-based), or the number of reads from that cluster that were present (Histogram-based. An example of this is illustrated as a part of Figure \ref{pipeline}. We implemented this feature extraction method in Python.

### 2016-08-22 latex/input/related.tex
updated results in paper, edited plots, wrote script to calculate AUC-ROC

An example of Vocabulary-based methods are Bag of Words (BoW) methods, which involve the following three-step process: (i) Cluster the instances to create classes of instances; (ii) for each bag, map the clusters of instances in that bag to a feature vector; and (iii) use a standard classifier that uses the feature vectors to predict group labels \cite{amores13}. Amores found the Distance-based Bag of Words (D-BoW) method to be the second most effective of all tested methods, and the most effective one that was also time-efficient \cite{amores13}. The distinguishing feature in D-BoW methods is that the values for the feature vector represent the instance that has the smallest distance to the cluster center. Histogram-based Bag of Words (H-BoW) methods count the number of instances in a given cluster instead of keeping track of the closest-matching instance to that cluster. H-BoW methods were found to be among the more effective methods, but not quite as good as D-BoW \cite{amores13}.

>>>>

An example of Vocabulary-based methods are Bag of Words (BoW) methods, which involve the following three-step process: (i) Cluster the instances to create classes of instances; (ii) for each bag, map the clusters of instances in that bag to a feature vector; and (iii) use a standard classifier that uses the feature vectors to predict group labels \cite{amores13}. Amores found the Distance-based Bag of Words (D-BoW) method to be the second most effective of all tested methods, and the most effective one that was also time-efficient (linear, rather than quadratic, in the number of bags and number of instances per bag) \cite{amores13}. The distinguishing feature in D-BoW methods is that the values for the feature vector represent the instance that has the smallest distance to the cluster center. Histogram-based Bag of Words (H-BoW) methods count the number of instances from each cluster there are in each bag, instead of keeping track of the closest-matching instance to that cluster. This intuitively has appeal in the domain of microbiome analysis, as the relative quantities of different species of bacteria is important. H-BoW methods were found by Amores to be somewhat less effective than D-BoW methods on average, but performed the best out of all algorithms on several datasets, indicating that this method performs very well on real problems that it is well-suited for \cite{amores13}. The D-BoW and H-BoW feature extraction methods are illustrated as part of Figure \ref{pipeline}.
